---
title: "Univariate Viz"
---

Use this file for practice with the **univariate viz** in-class activity.  Refer to the class website for details.




title: "Univariate Viz"
number-sections: true
execute: 
  eval: false


```{r}
library(tidyverse)
library(dplyr)

# Import data
hikes <- read.csv("https://mac-stat.github.io/data/high_peaks.csv")
```


## Exercises

### Exercise 1: Research Questions {.unnumbered}

a.  how many hikes are there in each category?

b.  What's the range in elevations?



### Exercise 3: Bar Chart of Ratings - Part 1 {.unnumbered}

```{r}
#| eval: true

ggplot(hikes, aes(x = rating))
```

-   just a blank canvas
-   name of the dataset
-   indicate which variable to plot on x-axis
-   `aes`thetics



### Exercise 4: Bar Chart of Ratings - Part 2 {.unnumbered}

```{r}
#| eval: true


ggplot(hikes, aes(x = rating)) +
  geom_bar()


ggplot(hikes, aes(x = rating)) +
  geom_bar() +
  labs(x = "Rating", y = "Number of hikes")


ggplot(hikes, aes(x = rating)) +
  geom_bar(fill = "blue") +
  labs(x = "Rating", y = "Number of hikes")


ggplot(hikes, aes(x = rating)) +
  geom_bar(color = "orange", fill = "blue") +
  labs(x = "Rating", y = "Number of hikes")


ggplot(hikes, aes(x = rating)) +
  geom_bar(color = "orange", fill = "blue")  +
  labs(x = "Rating", y = "Number of hikes") + 
  theme_minimal()
```


### Exercise 5: Bar Chart Follow-up {.unnumbered}

#### Part a {.unnumbered}

-   To indicate we're still adding layers to / modifying our plot.
-   Bars are the `geom`etric elements we're adding in this layer.
-   labels
-   `fill` fills in the bars. `color` outlines the bars.

#### Part b {.unnumbered}

Most hikes are moderate, the fewest number are difficult.

#### Part c {.unnumbered}

I don't like that the categories are alphabetical, not in order of difficulty level.



### Exercise 6: Sad Bar Chart {.unnumbered}

There are too many different outcomes of elevation.

```{r}
#| eval: true

ggplot(hikes, aes(x = elevation)) + 
  geom_bar()
```



### Exercise 7: A Histogram of Elevation {.unnumbered}

#### Part a {.unnumbered}

-   6
-   1 + 1 = 2

#### Part b {.unnumbered}

Elevations range from roughly 3700 to 5500 feet. Elevations vary from hike to hike relatively *normally* (with a bell shape) around a typical elevation of roughly 4500 feet.



### Exercise 9: Building Histograms - Part 2 {.unnumbered}

```{r}
#| eval: true

ggplot(hikes, aes(x = elevation)) +
  geom_histogram()

ggplot(hikes, aes(x = elevation)) +
  geom_histogram(color = "white") 

ggplot(hikes, aes(x = elevation)) +
  geom_histogram(color = "white", fill = "blue") 


ggplot(hikes, aes(x = elevation)) +
  geom_histogram(color = "white") +
  labs(x = "Elevation (feet)", y = "Number of hikes")


ggplot(hikes, aes(x = elevation)) +
  geom_histogram(color = "white", binwidth = 1000) +
  labs(x = "Elevation (feet)", y = "Number of hikes")


ggplot(hikes, aes(x = elevation)) +
  geom_histogram(color = "white", binwidth = 5) +
  labs(x = "Elevation (feet)", y = "Number of hikes")


ggplot(hikes, aes(x = elevation)) +
  geom_histogram(color = "white", binwidth = 200) +
  labs(x = "Elevation (feet)", y = "Number of hikes")
```


### Exercise 10: Histogram Follow-up {.unnumbered}

-   `geom_histogram()`
-   `color` outlined the bars and `fill` filled them
-   easier to distinguish between the bars
-   changed the bin width
-   we lump too many hikes together and lose track of the nuances
-   we don't lump enough hikes together and lose track of the bigger picture trends


### Exercise 11: Density plots {.unnumbered}

```{r}
#| eval: true

ggplot(hikes, aes(x = elevation)) +
 geom_density(color = "blue", fill = "orange")
```


